---
title: "Aquaculture summary"
subtitle: "What the Norwegian aquaculture site table contains"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

A summary of the 3,743 Norwegian marine aquaculture sites (see
[Aquaculture data](aquaculture-data.qmd) for how the table is built). The figures
below describe where the sites are, when they operated, how large they are, and what
they farm.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

con  <- dbConnect(SQLite(), "data/db/aquaculture_no.sqlite")
aqua <- dbReadTable(con, "aquaculture") |> as_tibble()
dbDisconnect(con)

aqua <- aqua |> mutate(status = ifelse(active == 1, "active", "closed"))
ACTIVE <- "#4477AA"; CLOSED <- "#CCBB44"
theme_set(theme_minimal(base_size = 12))
```

## Where the sites are

Every site as a point. Even without a base map, the points trace the Norwegian
coast, from the south around 58 degrees north to the far north beyond 71 degrees.
Active sites are shown against the closed ones.

```{r}
#| label: map
#| fig-cap: "Norwegian marine aquaculture sites, active and closed. Points are plotted at their latitude and longitude."
#| fig-width: 6
#| fig-height: 6.5
midlat <- mean(range(aqua$latitude))
ggplot(aqua |> arrange(active), aes(longitude, latitude, colour = status)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_colour_manual(values = c(active = ACTIVE, closed = CLOSED), name = NULL) +
  coord_fixed(ratio = 1 / cos(midlat * pi / 180)) +
  labs(x = "longitude", y = "latitude") +
  guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

Almost all sites are placed in the sea (3,450), with a smaller number of land-based
sites that use seawater (286). Three counties on the west and north coast, Vestland,
Nordland and Trøndelag, hold more than half of all sites.

```{r}
#| label: county
#| fig-cap: "Number of sites per county (top ten), active and closed."
#| fig-width: 6.5
#| fig-height: 3.6
top_counties <- aqua |> count(county, sort = TRUE) |> slice_head(n = 10) |> pull(county)
aqua |>
  filter(county %in% top_counties) |>
  count(county, status) |>
  ggplot(aes(n, reorder(county, n), fill = status)) +
  geom_col() +
  scale_fill_manual(values = c(active = ACTIVE, closed = CLOSED), name = NULL) +
  labs(x = "sites", y = NULL) +
  theme(panel.grid.major.y = element_blank(), legend.position = "top")
```

## When they operated

Sites are grouped by the decade they were cleared. Clearances rose through the 1990s
and 2000s; many early sites have since been withdrawn, so the closed share is largest
among the older decades.

```{r}
#| label: decade
#| fig-cap: "Sites by the decade they were cleared, active and closed."
#| fig-width: 6.5
#| fig-height: 3.4
aqua |>
  filter(!is.na(start_year)) |>
  mutate(decade = paste0((start_year %/% 10) * 10, "s")) |>
  count(decade, status) |>
  ggplot(aes(decade, n, fill = status)) +
  geom_col() +
  scale_fill_manual(values = c(active = ACTIVE, closed = CLOSED), name = NULL) +
  labs(x = NULL, y = "sites") +
  theme(panel.grid.major.x = element_blank(), legend.position = "top")
```

## How large they are

Most sites report their capacity as a permitted biomass in tonnes. Among active sites
with a tonnes capacity, the distribution is shown below (log scale, because a few
very large salmon sites reach tens of thousands of tonnes while most are far smaller).

```{r}
#| label: capacity
#| fig-cap: "Permitted capacity of active sites that report tonnes (log scale)."
#| fig-width: 6.5
#| fig-height: 3.2
aqua |>
  filter(active == 1, !is.na(capacity_tonnes), capacity_tonnes > 0) |>
  ggplot(aes(capacity_tonnes)) +
  geom_histogram(bins = 30, fill = ACTIVE) +
  scale_x_log10(labels = scales::comma) +
  labs(x = "permitted capacity (tonnes, log scale)", y = "active sites")
```

## What they farm

The register lists the species cleared at each site. A site can carry several. The
table counts how many sites list each of the main commercial groups (a site is
counted once per group it lists).

```{r}
#| label: species
#| fig-cap: "Number of sites listing each main species group (a site may appear in more than one)."
#| fig-width: 6.5
#| fig-height: 3.4
groups <- tibble::tribble(
  ~group,                 ~pattern,
  "Salmon",               "Atlantic salmon",
  "Trout / rainbow trout","trout|Trout",
  "Cod",                  "Atlantic cod",
  "Halibut",              "halibut",
  "Blue mussel",          "Blue mussel",
  "Scallop / oyster",     "scallop|oyster",
  "Kelp / seaweed",       "kelp|Dulse|wrack|laver|Sea lettuce",
  "Cleaner fish",         "wrasse|Lumpfish"
)
grp_counts <- groups |>
  rowwise() |>
  mutate(sites = sum(grepl(pattern, aqua$fish_types, ignore.case = TRUE), na.rm = TRUE)) |>
  ungroup()

grp_counts |>
  ggplot(aes(sites, reorder(group, sites))) +
  geom_col(fill = ACTIVE) +
  geom_text(aes(label = sites), hjust = -0.15, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = "sites", y = NULL) +
  theme(panel.grid.major.y = element_blank())
```

Salmon and trout dominate: roughly 2,000 sites list each. Blue mussel is the main
shellfish, and a growing number of sites list kelp and seaweed.

## Distance from sediment sites to the nearest farm

Each sediment site in Norway carries its distance to the nearest aquaculture site.
The table summarises that distance per source (Norwegian sites only). Vannmiljø, the
Norwegian coastal monitoring source, sits closest to the farms; Mareano's offshore
and Arctic stations are much farther away.

```{r}
#| label: distance
#| fig-cap: "Distance from each Norwegian sediment site to the nearest aquaculture site."
sources <- tibble::tribble(
  ~Source,      ~stem,
  "Mareano",    "mareano",
  "Vannmiljø",  "vannmiljo",
  "ICES-DOME",  "ices_dome",
  "MUDAB",      "mudab"
)
dist_summary <- purrr::pmap_dfr(sources, function(Source, stem) {
  con <- dbConnect(SQLite(), sprintf("data/db/%s_clean.sqlite", stem))
  d <- dbGetQuery(con, "SELECT dist_to_aquaculture d FROM site WHERE dist_to_aquaculture IS NOT NULL")$d
  dbDisconnect(con)
  tibble(Source = Source, `n sites` = length(d),
         `median km` = round(median(d), 1), `mean km` = round(mean(d), 1),
         `max km` = round(max(d)))
})
kable(dist_summary)
```

::: {.callout-note}
The aquaculture data are Norwegian only, so the distance column and this table cover
Norwegian sediment sites only. Capacity is the permitted (cleared) capacity, and only
the tonnes-based sites are summarised above, since area and count capacities are not
comparable.
:::
