---
title: "DB Schema (Clean)"
subtitle: "The tables and columns of the clean databases these analyses read"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
.mermaid { max-width: 620px; margin-left: auto; margin-right: auto; }
.mermaid svg { max-width: 100% !important; width: 100%; height: auto; }
```

The analyses on this site read the five clean databases, one per source. Each
`<source>_clean.sqlite` has the seven shared tables carried from the slim schema
(`element`, `dataset`, `site`, `event`, `subsample`, `measurement`, `method`), plus
`grain_size_fraction` where the source has grain-size (not 4Demon). How they are
built is documented on the
[slim site](https://seafood-hazards.github.io/multised-slim/); this page is a
reference for the table layout the analyses depend on.

## Schema diagram

The hierarchy flows from `dataset` and `site` through `event` and `subsample` down
to `measurement`. `element` is a lookup for both `method` and `measurement` via the
`symbol` key. The `grain_size_fraction` table (absent for 4Demon) hangs off
`subsample` alongside `measurement`.

```{mermaid}
erDiagram
  dataset   ||--o{ event               : "dataset_id"
  site      ||--o{ event               : "site_id"
  event     ||--o{ subsample           : "event_id"
  subsample ||--o{ measurement         : "subsample_id"
  subsample ||--o{ grain_size_fraction : "subsample_id"
  element   ||--o{ measurement         : "symbol"
  element   ||--o{ method              : "symbol"
  method    ||--o{ measurement         : "method_id"

  element {
    TEXT symbol PK
  }
  dataset {
    INTEGER dataset_id PK
  }
  site {
    INTEGER site_id PK
  }
  event {
    INTEGER event_id PK
    INTEGER dataset_id FK
    INTEGER site_id FK
  }
  subsample {
    INTEGER subsample_id PK
    INTEGER event_id FK
  }
  measurement {
    INTEGER measurement_id PK
    INTEGER subsample_id FK
    TEXT symbol FK
    INTEGER method_id FK
  }
  method {
    INTEGER method_id PK
    TEXT symbol FK
  }
  grain_size_fraction {
    INTEGER gsf_id PK
    INTEGER subsample_id FK
    TEXT symbol FK
  }
```

Each entity above shows only its keys; the full column list for every table is in
the sections below.

::: {.callout-note}
`measurement.symbol` is a denormalisation: it is derivable via
`method_id -> method.symbol` but kept as a key for query efficiency.
:::

```{r}
library(tibble)
library(knitr)
```

## dataset

One row per data-collection campaign (the provenance of a set of events).

```{r}
tribble(
  ~Column,          ~Type,     ~Meaning,
  "dataset_id",     "INTEGER", "Primary key.",
  "dataset_name",   "TEXT",    "Name of the data-collection campaign / dataset.",
  "dataset_code",   "TEXT",    "Source-native dataset code, where recorded.",
  "dataset_group",  "TEXT",    "Dataset grouping, where recorded.",
  "source",         "TEXT",    "Source database name (e.g. Mareano, ICES-DOME).",
  "source_type",    "TEXT",    "Kind of source (e.g. national, international).",
  "url",            "TEXT",    "Source URL, where recorded.",
  "doi",            "TEXT",    "Dataset DOI, where recorded.",
  "country",        "TEXT",    "Country of origin.",
  "region",         "TEXT",    "Marine region, where recorded.",
  "institute",      "TEXT",    "Institute responsible for the data.",
  "institute_code", "TEXT",    "Institute code, where recorded.",
  "accessed",       "TEXT",    "Date the source was accessed."
) |> kable()
```

## site

One row per geographic location, keyed (in slim) on latitude/longitude rounded to
three decimal places.

```{r}
tribble(
  ~Column,         ~Type,     ~Meaning,
  "site_id",       "INTEGER", "Primary key.",
  "latitude",      "REAL",    "Latitude in decimal degrees.",
  "longitude",     "REAL",    "Longitude in decimal degrees.",
  "depth",         "REAL",    "Water depth (m), where recorded.",
  "country",       "TEXT",    "Nearest country.",
  "country_code",  "TEXT",    "ISO country code.",
  "dist_to_coast", "REAL",    "Distance to nearest coastline (km).",
  "municipality",  "TEXT",    "Nearest municipality.",
  "sea_name",      "TEXT",    "Name of the sea or ocean body."
) |> kable()
```

## event

One row per sampling event (a single deployment of a gear at a site).

```{r}
tribble(
  ~Column,         ~Type,     ~Meaning,
  "event_id",      "INTEGER", "Primary key.",
  "dataset_id",    "INTEGER", "-> dataset.",
  "site_id",       "INTEGER", "-> site.",
  "year",          "INTEGER", "Year of the sampling event.",
  "date",          "TEXT",    "Date of the sampling event, where recorded.",
  "sampling_tool", "TEXT",    "Sampling gear, descriptive name.",
  "n_layers",      "INTEGER", "Number of subsamples (depth layers) under the event."
) |> kable()
```

`datetime` / `time` from the slim schema are dropped in the clean stage.

## subsample

One row per subsample (a depth slice of a sampling event), with depth, the
supporting-data flags, a target-fraction summary, and the mud content.

```{r}
tribble(
  ~Column,             ~Type,     ~Meaning,
  "subsample_id",      "INTEGER", "Primary key.",
  "event_id",          "INTEGER", "-> event.",
  "depth_from",        "REAL",    "Top of the slice, cm (NULL if the source's depth was removed or never recorded).",
  "depth_to",          "REAL",    "Bottom of the slice, cm (NULL if the source's depth was removed or never recorded).",
  "depth_flag",        "TEXT",    "'implausible' where the reported depth was out of range or inverted and was removed (Vannmiljø only); NULL otherwise. Lets a user spot and drop depth-less rows.",
  "fe_exist",          "INTEGER", "1/0: an Fe value exists for this subsample.",
  "al_exist",          "INTEGER", "1/0: an Al value exists.",
  "org_exist",         "INTEGER", "1/0: an organic carbon value exists.",
  "comp_exist",        "INTEGER", "1/0: grain-size composition exists.",
  "target_frac_class", "TEXT",    "Fraction summary of the subsample's target measurements: 'bulk', 'sieved', or 'mixed' (targets span both). NULL if no target-metal measurement (grain-size-only, or normaliser/organic-only subsamples).",
  "target_sieve_um",   "REAL",    "Target sieve cutoff where unambiguous, else NULL.",
  "fines_lt63",        "REAL",    "% finer than 63 um (clay + silt mud content), else NULL.",
  "fines_basis",       "TEXT",    "How fines_lt63 was derived."
) |> kable()
```

The authoritative bulk/sieved value is per measurement (`measurement.frac_class`);
`subsample.target_frac_class` is a convenience summary over the targets, with
`mixed` marking the (few) subsamples whose target metals span both fractions.

## measurement

One row per cleaned chemistry measurement: the 7 target elements, the Fe/Al
normalisers, and organic carbon (CORG / TOC*), told apart by `element.category`.
Failed-QC rows are gone; duplicates and technical replicates are collapsed.
Grain-size composition has moved to `grain_size_fraction`.

```{r}
tribble(
  ~Column,        ~Type,    ~Meaning,
  "measurement_id","INTEGER","Primary key (carried from slim).",
  "subsample_id",  "INTEGER","-> subsample.",
  "symbol",        "TEXT",   "Element, ICES name (-> element).",
  "value",         "REAL",   "Value in the original (ICES-named) unit.",
  "unit",          "TEXT",   "Original reporting unit.",
  "value_std",     "REAL",   "Standardised value, mg/kg (mean where replicates were averaged).",
  "unit_std",      "TEXT",   "Always mg/kg.",
  "value_sd",      "REAL",   "SD across technical replicates, else NULL.",
  "n_rep",         "INTEGER","Number of distinct replicate values averaged (1 = single).",
  "value_uncrt",   "REAL",   "Analytical uncertainty, mg/kg (ICES-DOME only), else NULL.",
  "matrix",        "TEXT",   "Raw ICES sediment-fraction code (provenance), where recorded.",
  "frac_class",    "TEXT",   "Sediment fraction measured: 'bulk' (whole sample or the under-2 mm fraction) or 'sieved' (a fine sub-fraction).",
  "sieve_um",      "REAL",   "Sieve cutoff in um for sieved rows (e.g. 63, 20), NULL for bulk.",
  "method_id",     "INTEGER","-> method."
) |> kable()
```

`frac_class` / `sieve_um` are derived from `matrix` (kept alongside). Bulk versus
sieved is recorded per measurement because it varies within a subsample: a sample's
metals and its organic carbon, and even different target metals, are often measured
on different fractions.

## element

Lookup mapping each chemical symbol to its name, measurand category, and CAS
number. Referenced by both `measurement` and `method` via `symbol`.

```{r}
tribble(
  ~Column,    ~Type,  ~Meaning,
  "symbol",   "TEXT", "Primary key: element in the ICES name (e.g. CO, CU, CORG).",
  "name",     "TEXT", "Full element / measurand name.",
  "category", "TEXT", "Measurand class: target / reference / organic / composition.",
  "cas",      "TEXT", "CAS registry number, where known."
) |> kable()
```

## method

One row per distinct analytical method (technique, lab, and limits) for a symbol.

```{r}
tribble(
  ~Column,              ~Type,     ~Meaning,
  "method_id",          "INTEGER", "Primary key.",
  "symbol",             "TEXT",    "-> element.",
  "method",             "TEXT",    "Analytical technique (e.g. ICP-MS, ICP-AES).",
  "method_description", "TEXT",    "Free-text method description, where recorded.",
  "lab",                "TEXT",    "Laboratory that performed the analysis, where recorded.",
  "lab_name",           "TEXT",    "Full laboratory name, where recorded.",
  "lod",                "REAL",    "Limit of detection, where recorded (Mareano's lld maps here).",
  "loq",                "REAL",    "Limit of quantification, where recorded.",
  "limit_unit",         "TEXT",    "Reporting unit of lod / loq."
) |> kable()
```

All seven shared tables are identical in structure across the five sources; values
are absent (NULL) where a source does not record a column.

## grain_size_fraction (per fraction measurement)

The grain-size distribution moved out of `measurement`. One row per grain-size
mass-fraction value. Absent for 4Demon.

```{r}
tribble(
  ~Column,       ~Type,    ~Meaning,
  "gsf_id",      "INTEGER","Primary key (surrogate; the subsample_id/symbol/matrix combination is not unique).",
  "subsample_id","INTEGER","-> subsample.",
  "symbol",      "TEXT",   "The source's grain-size code.",
  "matrix",      "TEXT",   "Sample fraction the value is a percentage of (where recorded).",
  "lo_um",       "REAL",   "Lower size bound, um.",
  "hi_um",       "REAL",   "Upper size bound, um (NULL = open, e.g. gravel).",
  "value_pct",   "REAL",   "Corrected mass fraction, %."
) |> kable()
```

## What changed from slim

```{r}
tribble(
  ~Change,                              ~Detail,
  "Rows removed",                       "Out of range, invalid, below limit, wet weight, source-specific QC.",
  "Duplicates / replicates collapsed",  "One value each; replicates gain value_sd + n_rep.",
  "Grain-size out of measurement",      "Grain-size composition moves to grain_size_fraction; all chemistry (incl. organic carbon) stays in measurement.",
  "Sediment fraction as frac_class",    "matrix gains frac_class (bulk/sieved) + sieve_um; a target bulk/sieved/mixed summary is added to subsample.",
  "Flag columns dropped",               "The slim QC flags were consumed by the cleaning and are not carried.",
  "Depths in cm",                       "ICES-DOME converted from metres; Vannmiljo corrupt depths nulled.",
  "Symbols / units unified",            "ICES names; TOC -> CORG; value_std in mg/kg; CAS added to element."
) |> kable()
```
