---
title: "DB Schema (Merged)"
subtitle: "The tables and columns of the single merged database"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
.mermaid { max-width: 620px; margin-left: auto; margin-right: auto; }
.mermaid svg { max-width: 100% !important; width: 100%; height: auto; }
```

The **merged** database (`multised_merged.sqlite`) combines the five clean databases
into one, with cross-source duplicates removed. It keeps the same eight tables and
the same relationships as the [clean schema](./db-schema-clean.qmd); this page lists
only what the merge **adds**. How it is built is on the merged-database pages
([union](./merge-union.qmd), [deduplication](./merge-dedup.qmd),
[finalise and retention](./merge-retention.qmd)).

## What the merge adds

Every table except the shared `element` lookup gains two provenance columns, and the
`measurement` table gains a harmonised sieve cutoff:

- **`source`** (TEXT): which source database the row came from (Mareano, 4Demon,
  MUDAB, Vannmiljø, ICES-DOME).
- **`src_<table>_id`** (TEXT): the row's original source-prefixed key (e.g.
  `ice_107870`), kept so a merged row can be traced back to its clean database.
- **`sieve_um_std`** / **`sieve_class`** on `measurement`: a canonical sieve cutoff
  (62 folded to 63) and its label (`<63um`, `<20um`), so like-cutoff sieved data can
  be compared. NULL for bulk.

The primary keys are renumbered to clean contiguous integers across the whole merged
database (the source-prefixed keys survive only in the `src_*_id` columns). Bulk and
sieved are kept as separate tracks, as in the clean stage.

## Schema diagram

The structure is identical to the clean schema.

```{mermaid}
erDiagram
  dataset   ||--o{ event               : "dataset_id"
  site      ||--o{ event               : "site_id"
  event     ||--o{ subsample           : "event_id"
  subsample ||--o{ measurement         : "subsample_id"
  subsample ||--o{ grain_size_fraction : "subsample_id"
  element   ||--o{ measurement         : "symbol"
  element   ||--o{ method              : "symbol"
  method    ||--o{ measurement         : "method_id"

  element {
    TEXT symbol PK
  }
  dataset {
    INTEGER dataset_id PK
    TEXT source
  }
  site {
    INTEGER site_id PK
    TEXT source
  }
  event {
    INTEGER event_id PK
    INTEGER dataset_id FK
    INTEGER site_id FK
    TEXT source
  }
  subsample {
    INTEGER subsample_id PK
    INTEGER event_id FK
    TEXT source
  }
  measurement {
    INTEGER measurement_id PK
    INTEGER subsample_id FK
    TEXT symbol FK
    INTEGER method_id FK
    TEXT source
  }
  method {
    INTEGER method_id PK
    TEXT symbol FK
    TEXT source
  }
  grain_size_fraction {
    INTEGER gsf_id PK
    INTEGER subsample_id FK
    TEXT symbol FK
    TEXT source
  }
```

```{r}
library(tibble)
library(knitr)
```

## dataset

One row per data-collection campaign. Clean columns plus `source` and provenance.

```{r}
tribble(
  ~Column,          ~Type,     ~Meaning,
  "dataset_id",     "INTEGER", "Primary key (renumbered across all sources).",
  "dataset_name",   "TEXT",    "Name of the data-collection campaign / dataset.",
  "dataset_code",   "TEXT",    "Source-native dataset code, where recorded.",
  "dataset_group",  "TEXT",    "Dataset grouping, where recorded.",
  "source",         "TEXT",    "Source database name.",
  "source_type",    "TEXT",    "Kind of source (e.g. national, international).",
  "url",            "TEXT",    "Source URL, where recorded.",
  "doi",            "TEXT",    "Dataset DOI, where recorded.",
  "country",        "TEXT",    "Country of origin.",
  "region",         "TEXT",    "Marine region, where recorded.",
  "institute",      "TEXT",    "Institute responsible for the data.",
  "institute_code", "TEXT",    "Institute code, where recorded.",
  "accessed",       "TEXT",    "Date the source was accessed.",
  "src_dataset_id", "TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

## site

One row per geographic location. Carries the geo-enrichment (including
`dist_to_aquaculture`) plus `source` and provenance.

```{r}
tribble(
  ~Column,               ~Type,     ~Meaning,
  "site_id",             "INTEGER", "Primary key (renumbered across all sources).",
  "latitude",            "REAL",    "Latitude in decimal degrees.",
  "longitude",           "REAL",    "Longitude in decimal degrees.",
  "depth",               "REAL",    "Water depth (m), where recorded.",
  "country",             "TEXT",    "Nearest country.",
  "country_code",        "TEXT",    "ISO country code.",
  "dist_to_coast",       "REAL",    "Distance to nearest coastline (km).",
  "municipality",        "TEXT",    "Nearest municipality.",
  "sea_name",            "TEXT",    "Name of the sea or ocean body.",
  "dist_to_aquaculture", "REAL",    "Distance to nearest Norwegian aquaculture site (km); Norway sites only, else NULL.",
  "source",              "TEXT",    "Source database name.",
  "src_site_id",         "TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

## event

One row per sampling event.

```{r}
tribble(
  ~Column,         ~Type,     ~Meaning,
  "event_id",      "INTEGER", "Primary key (renumbered across all sources).",
  "dataset_id",    "INTEGER", "-> dataset.",
  "site_id",       "INTEGER", "-> site.",
  "year",          "INTEGER", "Year of the sampling event.",
  "date",          "TEXT",    "Date of the sampling event, where recorded.",
  "sampling_tool", "TEXT",    "Sampling gear, descriptive name.",
  "n_layers",      "INTEGER", "Number of subsamples (depth layers) under the event.",
  "source",        "TEXT",    "Source database name.",
  "src_event_id",  "TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

## subsample

One row per subsample (a depth slice of an event).

```{r}
tribble(
  ~Column,             ~Type,     ~Meaning,
  "subsample_id",      "INTEGER", "Primary key (renumbered across all sources).",
  "event_id",          "INTEGER", "-> event.",
  "depth_from",        "REAL",    "Top of the slice, cm (NULL if removed or never recorded).",
  "depth_to",          "REAL",    "Bottom of the slice, cm (NULL if removed or never recorded).",
  "depth_flag",        "TEXT",    "'implausible' where the reported depth was removed (Vannmiljø only), else NULL.",
  "fe_exist",          "INTEGER", "1/0: an Fe value exists for this subsample.",
  "al_exist",          "INTEGER", "1/0: an Al value exists.",
  "org_exist",         "INTEGER", "1/0: an organic carbon value exists.",
  "comp_exist",        "INTEGER", "1/0: grain-size composition exists.",
  "target_frac_class", "TEXT",    "Target-fraction summary: 'bulk', 'sieved', or 'mixed', else NULL.",
  "target_sieve_um",   "REAL",    "Target sieve cutoff where unambiguous, else NULL.",
  "fines_lt63",        "REAL",    "% finer than 63 um (mud content), else NULL.",
  "fines_basis",       "TEXT",    "How fines_lt63 was derived.",
  "source",            "TEXT",    "Source database name.",
  "src_subsample_id",  "TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

## measurement

One row per cleaned chemistry measurement, after cross-source duplicates are removed.
Carries the harmonised sieve cutoff and provenance.

```{r}
tribble(
  ~Column,             ~Type,     ~Meaning,
  "measurement_id",    "INTEGER", "Primary key (renumbered across all sources).",
  "subsample_id",      "INTEGER", "-> subsample.",
  "symbol",            "TEXT",    "Element, ICES name (-> element).",
  "value",             "REAL",    "Value in the original (ICES-named) unit.",
  "unit",              "TEXT",    "Original reporting unit.",
  "value_std",         "REAL",    "Standardised value, mg/kg.",
  "unit_std",          "TEXT",    "Always mg/kg.",
  "value_sd",          "REAL",    "SD across technical replicates, else NULL.",
  "n_rep",             "INTEGER", "Number of replicate values averaged (1 = single).",
  "value_uncrt",       "REAL",    "Analytical uncertainty, mg/kg (ICES-DOME only), else NULL.",
  "matrix",            "TEXT",    "Raw ICES sediment-fraction code (provenance).",
  "frac_class",        "TEXT",    "Sediment fraction: 'bulk' or 'sieved'.",
  "sieve_um",          "REAL",    "Sieve cutoff in um for sieved rows, NULL for bulk.",
  "method_id",         "INTEGER", "-> method.",
  "sieve_um_std",      "REAL",    "Harmonised sieve cutoff (62 folded to 63), NULL for bulk.",
  "sieve_class",       "TEXT",    "Sieve label, e.g. '<63um' / '<20um', NULL for bulk.",
  "source",            "TEXT",    "Source database name.",
  "src_measurement_id","TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

## element

The shared element lookup. Unchanged from clean: a single vocabulary across all
sources, so it carries **no** `source` column.

```{r}
tribble(
  ~Column,    ~Type,  ~Meaning,
  "symbol",   "TEXT", "Primary key: element in the ICES name (e.g. CO, CU, CORG).",
  "name",     "TEXT", "Full element / measurand name.",
  "category", "TEXT", "Measurand class: target / reference / organic / composition.",
  "cas",      "TEXT", "CAS registry number, where known."
) |> kable()
```

## method

One row per distinct analytical method.

```{r}
tribble(
  ~Column,              ~Type,     ~Meaning,
  "method_id",          "INTEGER", "Primary key (renumbered across all sources).",
  "symbol",             "TEXT",    "-> element.",
  "method",             "TEXT",    "Analytical technique (e.g. ICP-MS, ICP-AES).",
  "method_description", "TEXT",    "Free-text method description, where recorded.",
  "lab",                "TEXT",    "Laboratory that performed the analysis, where recorded.",
  "lab_name",           "TEXT",    "Full laboratory name, where recorded.",
  "lod",                "REAL",    "Limit of detection, where recorded.",
  "loq",                "REAL",    "Limit of quantification, where recorded.",
  "limit_unit",         "TEXT",    "Reporting unit of lod / loq.",
  "source",             "TEXT",    "Source database name.",
  "src_method_id",      "TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

## grain_size_fraction (per fraction measurement)

One row per grain-size mass-fraction value (4Demon contributes none).

```{r}
tribble(
  ~Column,       ~Type,     ~Meaning,
  "gsf_id",      "INTEGER", "Primary key (renumbered across all sources).",
  "subsample_id","INTEGER", "-> subsample.",
  "symbol",      "TEXT",    "The source's grain-size code.",
  "matrix",      "TEXT",    "Sample fraction the value is a percentage of.",
  "lo_um",       "REAL",    "Lower size bound, um.",
  "hi_um",       "REAL",    "Upper size bound, um (NULL = open, e.g. gravel).",
  "value_pct",   "REAL",    "Corrected mass fraction, %.",
  "source",      "TEXT",    "Source database name.",
  "src_gsf_id",  "TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

::: {.callout-note}
The merged database is one row per reading: a measurement reported by more than one
source is kept once, from the most preferred source (Mareano > 4Demon > MUDAB >
Vannmiljø > ICES-DOME). The `source` and `src_*_id` columns record where each
surviving row came from.
:::
