---
title: "Grain size"
subtitle: "Sediment fraction and the <63 µm fines behind the target-element concentrations"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Trace-element concentrations in sediment depend strongly on grain size: finer
material concentrates metals, so a **sieved <63 µm** analysis is not comparable
to a **bulk** (whole-sample) one. Following the EFSA data specification, this page
characterises, **separately per source**, how the target measurements split
between bulk and sieved, how much fine (<63 µm) material the samples carry, and how
concentration tracks grain size. All depths are kept but reported by the EFSA depth
bands (0 to 5, 5 to 40, over 40 cm), assigned from the subsample interval midpoint.

The analysis itself is run in the `sedimenter` pipeline
(`R/analysis/grainsize/`); this page renders the summary tables and figures from
its outputs.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

gdir <- "data/analysis/grainsize"
frac  <- read_csv(file.path(gdir, "grainsize_fraction_summary.csv"), show_col_types = FALSE)
fines <- read_csv(file.path(gdir, "grainsize_fines_summary.csv"),    show_col_types = FALSE)
concf <- read_csv(file.path(gdir, "grainsize_conc_vs_fines.csv"),    show_col_types = FALSE)
bvs   <- read_csv(file.path(gdir, "grainsize_bulk_vs_sieved.csv"),   show_col_types = FALSE)
pts   <- read_csv(file.path(gdir, "grainsize_targets_fines.csv"),    show_col_types = FALSE)

src_levels  <- c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon")
band_levels <- c("0-5", "5-40", ">40", "unknown")
ord_src  <- function(x) factor(x, levels = src_levels)
theme_set(theme_minimal(base_size = 12))
```

## 1. Sample fraction: bulk vs sieved

How the target measurements divide between bulk (whole sample or the under-2 mm
fraction) and sieved (a fine sub-fraction). Mareano and Vannmiljø report only bulk;
MUDAB and 4Demon are dominated by sieved; ICES-DOME carries both.

```{r}
#| label: fraction-table
frac |>
  group_by(Source, frac_class) |>
  summarise(n = sum(n), .groups = "drop") |>
  pivot_wider(names_from = frac_class, values_from = n, values_fill = 0) |>
  mutate(Source = ord_src(Source),
         total = bulk + sieved,
         `% sieved` = round(100 * sieved / total)) |>
  arrange(Source) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: fraction-fig
#| fig-cap: "Share of target measurements analysed on bulk versus sieved sediment, per source."
#| fig-width: 7
#| fig-height: 3.2
frac |>
  group_by(Source, frac_class) |>
  summarise(n = sum(n), .groups = "drop") |>
  group_by(Source) |>
  mutate(share = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(ord_src(Source), share, fill = frac_class)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(bulk = "#4477AA", sieved = "#CCBB44")) +
  labs(x = NULL, y = "share of measurements", fill = NULL) +
  coord_flip()
```

## 2. The <63 µm fines

The percentage of material finer than 63 µm (the clay + silt mud), per source and
depth band. Only the grain-size sources carry it (4Demon has none). `mud_dominated`
counts subsamples at least 50 % fines.

```{r}
#| label: fines-table
fines |>
  mutate(Source = ord_src(Source),
         band = factor(band, levels = band_levels)) |>
  arrange(Source, band) |>
  select(Source, band, n_subsamples, n_with_fines,
         mean, median, p25, p75, mud_dominated) |>
  kable(format.args = list(big.mark = ","))
```

## 3. Bulk versus sieved concentration

For the sources that carry both fractions (ICES-DOME, MUDAB), the sieved <63 µm
concentration sits well above the bulk one for the grain-size-controlled metals
(CO, CU, MN, ZN), typically two to three times higher. Molybdenum is the exception
(redox / organic controlled, not grain-size controlled), so its sieved values are
not elevated.

```{r}
#| label: bvs-table
bvs |>
  filter(Source %in% c("ICES-DOME", "MUDAB")) |>
  select(Source, symbol, frac_class, n, median) |>
  pivot_wider(names_from = frac_class, values_from = c(n, median)) |>
  transmute(Source = ord_src(Source), Element = symbol,
            `n bulk` = n_bulk, `n sieved` = n_sieved,
            `median bulk` = median_bulk, `median sieved` = median_sieved,
            `sieved / bulk` = round(median_sieved / median_bulk, 1)) |>
  arrange(Source, Element) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: bvs-fig
#| fig-cap: "Median target concentration by fraction, for the sources carrying both (log scale)."
#| fig-width: 7.5
#| fig-height: 3.6
bvs |>
  filter(Source %in% c("ICES-DOME", "MUDAB"),
         symbol %in% c("CO", "CU", "MN", "MO", "ZN")) |>
  ggplot(aes(symbol, median, fill = frac_class)) +
  geom_col(position = position_dodge()) +
  scale_y_log10() +
  scale_fill_manual(values = c(bulk = "#4477AA", sieved = "#CCBB44")) +
  facet_wrap(~ Source) +
  labs(x = NULL, y = "median concentration (mg/kg, log)", fill = NULL)
```

## 4. The grain-size effect (bulk samples)

Within **bulk** samples, where the reported value describes the whole sample,
concentration rises with the fines content, as expected. The table gives the
Spearman correlation of concentration with `fines_lt63` per element and band
(bulk only, at least 10 samples); the figure shows it for Mareano, whose bulk,
clean grain-size gives the clearest signal.

```{r}
#| label: concf-table
concf |>
  filter(band %in% c("0-5", "5-40")) |>
  select(Source, symbol, band, n, rho) |>
  pivot_wider(names_from = band, values_from = c(n, rho)) |>
  transmute(Source = ord_src(Source), Element = symbol,
            `n (0-5)` = `n_0-5`, `rho (0-5)` = `rho_0-5`,
            `n (5-40)` = `n_5-40`, `rho (5-40)` = `rho_5-40`) |>
  arrange(Source, Element) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: concf-fig
#| fig-cap: "Mareano bulk target concentrations against the <63 µm fines content (log y, linear fit)."
#| fig-width: 7.5
#| fig-height: 3
pts |>
  filter(Source == "Mareano", frac_class == "bulk",
         symbol %in% c("CO", "CU", "ZN"),
         !is.na(fines_lt63), value_std > 0) |>
  ggplot(aes(fines_lt63, value_std)) +
  geom_point(alpha = 0.15, size = 0.6, colour = "#4477AA") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#AA3377", se = FALSE) +
  scale_y_log10() +
  facet_wrap(~ symbol, scales = "free_y") +
  labs(x = "fines <63 µm (%)", y = "concentration (mg/kg, log)")
```

::: {.callout-note}
The grain-size effect is measured on bulk samples only: a sieved <63 µm value is of
the fine fraction, not the whole sample, so relating it to the whole-sample fines
would not be meaningful. Molybdenum, selenium and iodine are kept in the tables but
are noisier (low counts, and not purely grain-size controlled).
:::
