---
title: "multised (clean analyses)"
subtitle: "Analyses on the QC-passed marine-sediment databases"
---

**multised** assembles a database of marine-sediment trace-element measurements
from five external sources, harmonises them into a common schema, and
progressively quality-controls the result. This site documents the final stage:
the **analyses performed on** the clean (QC-passed) databases, ending in a single
merged database for the background / pristine-sediment work.

How the clean databases are built (harmonisation, cleaning, and the grain-size
split) is documented separately on the
[slim site](https://seafood-hazards.github.io/multised-slim/), under **Clean
databases**.

## Data sources

```{r}
library(tibble)
library(knitr)

sources_tbl <- tribble(
  ~Source,      ~Origin,          ~Notes,
  "Mareano",    "Norway / IMR",   "Norwegian cruise data.",
  "Vannmiljø",  "Norway",         "Norwegian environmental monitoring.",
  "ICES-DOME",  "International",   "Rich sediment-composition data.",
  "MUDAB",      "Germany",        "German marine environmental database.",
  "4Demon",     "Belgium",        "Compiled historical Baltic / North Sea data (Belgian)."
)

kable(sources_tbl)
```

## Pipeline generations

The data moves through four generations, each one SQLite database **per source**
until the final merge:

1. **pilot**: each source's raw files parsed into a per-source database whose
   table structure follows that source. Documented on the individual pilot sites.
2. **slim**: every source reshaped into the same seven-table schema, then
   annotated with quality-control and marking flags (values untouched, suspicious
   records flagged rather than deleted). Documented on the
   [slim site](https://seafood-hazards.github.io/multised-slim/).
3. **clean** *(the input to this site)*: the flags are applied (uncertain rows
   removed), units are harmonised, and grain-size is split into its own table.
   Five per-source `<source>_clean.sqlite` databases.
4. **merged** *(the goal of this site)*: the five clean databases are combined
   into a single database for the background / pristine-sediment analysis.

## What this site covers

The analyses read the five clean databases and build towards the merged database.
The pages are grouped in the navigation bar:

**DB Design**

- [DB Schema (Clean)](./db-schema-clean.qmd): the table layout of the clean
  databases these analyses read, with the entity-relationship diagram.
- [Database Downloads](./database-downloads.qmd): download links for the five
  clean SQLite databases.

**Analyses** (added as each is completed)

```{r}
tribble(
  ~Analysis,                                      ~Status,
  "Iron / aluminium normalisation",               "planned",
  "Grain-size distribution analysis",             "planned",
  "Cross-source merge into one database",         "planned",
  "Background and pristine sediment definition",  "planned"
) |> kable()
```

::: {.callout-note}
Results will be added here as each analysis is completed. The analysis pages read
the clean databases live at render time.
:::
