---
title: "Merge step 2: deduplication"
subtitle: "Removing the same reading reported by more than one source"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The sources overlap. In particular ICES-DOME is an international archive that
re-hosts national data, so the same measurement can appear both in a national source
and again in ICES-DOME. The second step finds those duplicates and keeps only one copy.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

mdir <- "data/analysis/merge"
dd  <- read_csv(file.path(mdir, "merge_dedup.csv"), show_col_types = FALSE)
BLUE <- "#4477AA"
theme_set(theme_minimal(base_size = 12))
total_dup <- sum(dd$n)
```

## When two measurements are the same

Two measurements are treated as the same reading when they share **all** of:

- the same location (latitude and longitude rounded to three decimals),
- the **exact** sampling date (readings with no date are never merged),
- the same depth layer,
- the same element and the same track (bulk or sieved, and the sieve cutoff),
- and a standardised value within **1 percent** of each other.

This is deliberately strict: it targets re-hosted copies, not merely similar samples.
When a group of duplicates is found, the copy from the **most preferred source** is
kept and the others are dropped. The preference order is:

**Mareano > 4Demon > MUDAB > Vannmiljø > ICES-DOME.**

ICES-DOME is last because it is the aggregator; a national source is preferred over
its re-hosted copy.

## What was removed

In total **`r scales::comma(total_dup)`** measurements were flagged as duplicates
(about 8 percent). Almost all are ICES-DOME copies superseded by a national source,
and the largest single overlap is with MUDAB (the German data).

```{r}
#| label: winner-loser
#| fig-cap: "Duplicates removed, by which source superseded which. Almost all are ICES-DOME copies kept instead from a national source."
#| fig-width: 6.5
#| fig-height: 2.8
dd |>
  group_by(winner_source, loser_source) |>
  summarise(n = sum(n), .groups = "drop") |>
  mutate(pair = paste(winner_source, "over", loser_source)) |>
  ggplot(aes(n, reorder(pair, n))) +
  geom_col(fill = BLUE) +
  geom_text(aes(label = scales::comma(n)), hjust = -0.1, size = 3) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(x = "measurements removed", y = NULL) +
  theme(panel.grid.major.y = element_blank())
```

Broken down by element, the duplicates fall where the overlapping sources measure
most: copper, zinc, organic carbon and aluminium.

```{r}
#| label: by-element
#| fig-cap: "Duplicates removed per element."
#| fig-width: 6.5
#| fig-height: 3
dd |>
  group_by(symbol) |>
  summarise(n = sum(n), .groups = "drop") |>
  ggplot(aes(n, reorder(symbol, n))) +
  geom_col(fill = BLUE) +
  geom_text(aes(label = scales::comma(n)), hjust = -0.1, size = 3) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(x = "measurements removed", y = NULL) +
  theme(panel.grid.major.y = element_blank())
```

```{r}
#| label: track-table
dd |>
  group_by(`track` = frac_class) |>
  summarise(`measurements removed` = sum(n), .groups = "drop") |>
  kable()
```

::: {.callout-note}
Bulk and sieved are matched separately (the track is part of the key), so a bulk
reading is never removed in favour of a sieved one. The 1 percent, exact-date rule is
conservative: genuinely different samples at the same place are kept apart.
:::
