---
title: "Merge step 2: deduplication"
subtitle: "Removing the same reading reported by more than one source"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The sources overlap. ICES-DOME is an international archive that re-hosts national
data, so the same measurement can appear both in a national source and again in
ICES-DOME; and one Vannmiljø dataset is itself re-hosted Mareano data. The second
step finds those duplicates and keeps only one copy.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

mdir <- "data/analysis/merge"
dd  <- read_csv(file.path(mdir, "merge_dedup.csv"), show_col_types = FALSE)
BLUE <- "#4477AA"
theme_set(theme_minimal(base_size = 12))
total_dup <- sum(dd$n)
```

## When two measurements are the same

Two rules identify duplicates. Both keep the copy from the **most preferred source**
and drop the others; the preference order is:

**Mareano > 4Demon > MUDAB > Vannmiljø > ICES-DOME.**

ICES-DOME is last because it is the aggregator; a national source is preferred over
its re-hosted copy.

**The general rule.** Two measurements are the same reading when they share **all** of:

- the same location (latitude and longitude rounded to three decimals),
- the same sampling **year** (readings with no year are never merged),
- the same depth layer,
- the same element and the same track (bulk or sieved, and the sieve cutoff),
- and a standardised value within **1 percent** of each other.

The year (rather than the exact date) is used because a re-hosting archive often
rewrites the date field, so requiring an identical date would split copies that are
plainly the same reading. The strict 1 percent value match is what keeps genuinely
different samples apart, so relaxing the date to the year is safe: in practice it fires
almost only on real re-hosted overlaps, chiefly ICES-DOME copies of the German data
also held in MUDAB.

**The provenance rule.** One Vannmiljø dataset is literally named as re-hosted Mareano
data. Re-hosting nudged its values slightly, past the 1 percent match, so the general
rule misses it. For that dataset alone, a reading is a duplicate when native Mareano
has the same location, year, element and track within **5 percent**. The looser
tolerance is safe because the shared origin is known from the dataset name; the
Vannmiljø readings that native Mareano does not carry (it covers a shorter span of
years) are kept.

## What was removed

In total **`r scales::comma(total_dup)`** measurements were flagged as duplicates
(nearly 10 percent). Most are ICES-DOME copies superseded by a national source, and
the largest single overlap is with MUDAB (the German data); a smaller set is
Vannmiljø's re-hosted Mareano readings, superseded by native Mareano.

```{r}
#| label: winner-loser
#| fig-cap: "Duplicates removed, by which source superseded which. Almost all are ICES-DOME copies kept instead from a national source."
#| fig-width: 6.5
#| fig-height: 2.8
dd |>
  group_by(winner_source, loser_source) |>
  summarise(n = sum(n), .groups = "drop") |>
  mutate(pair = paste(winner_source, "over", loser_source)) |>
  ggplot(aes(n, reorder(pair, n))) +
  geom_col(fill = BLUE) +
  geom_text(aes(label = scales::comma(n)), hjust = -0.1, size = 3) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(x = "measurements removed", y = NULL) +
  theme(panel.grid.major.y = element_blank())
```

Broken down by element, the duplicates fall where the overlapping sources measure
most: copper, zinc, organic carbon and aluminium.

```{r}
#| label: by-element
#| fig-cap: "Duplicates removed per element."
#| fig-width: 6.5
#| fig-height: 3
dd |>
  group_by(symbol) |>
  summarise(n = sum(n), .groups = "drop") |>
  ggplot(aes(n, reorder(symbol, n))) +
  geom_col(fill = BLUE) +
  geom_text(aes(label = scales::comma(n)), hjust = -0.1, size = 3) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(x = "measurements removed", y = NULL) +
  theme(panel.grid.major.y = element_blank())
```

```{r}
#| label: track-table
dd |>
  group_by(`track` = frac_class) |>
  summarise(`measurements removed` = sum(n), .groups = "drop") |>
  kable()
```

::: {.callout-note}
Bulk and sieved are matched separately (the track is part of the key), so a bulk
reading is never removed in favour of a sieved one. The 1 percent value match is the
real guard on both rules: genuinely different samples at the same place and year are
kept apart.
:::
