---
title: "Merge step 1: union"
subtitle: "Combining the five clean databases into one"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The merged database is built in three steps. This page covers the first: bringing
the five per-source clean databases together into one, without yet removing anything.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

mdir <- "data/analysis/merge"
ret <- read_csv(file.path(mdir, "merge_retention.csv"),    show_col_types = FALSE)
tot <- read_csv(file.path(mdir, "merge_stage_totals.csv"), show_col_types = FALSE)

src_levels <- c("Mareano", "4Demon", "MUDAB", "Vannmiljø", "ICES-DOME")
BLUE <- "#4477AA"
theme_set(theme_minimal(base_size = 12))
```

## What the union does

Each source keeps its own primary keys (site 1, site 2, and so on), so simply
stacking the tables would collide them. To prevent that, **every key is prefixed with
a source code** while merging: `mar` for Mareano, `dem` for 4Demon, `mud` for MUDAB,
`van` for Vannmiljø, `ice` for ICES-DOME. A site becomes `mar_1`, `ice_1`, and the two
can never be confused. Every row also gains a `source` column. These prefixes are
temporary: the final step renumbers the keys to clean integers and keeps the source
and original id as provenance.

The element list is the one thing shared across sources (the same ICES element
symbols), so it is collapsed to a single vocabulary rather than repeated per source.

## What each source brings

At the union stage nothing is dropped, so the measurement counts are exactly those of
the clean databases. Vannmiljø and ICES-DOME are the largest contributors.

```{r}
#| label: contribution
#| fig-cap: "Measurements contributed by each source at the union stage (before deduplication)."
#| fig-width: 6.5
#| fig-height: 3.2
ret |>
  filter(table == "measurement") |>
  mutate(Source = factor(Source, levels = src_levels)) |>
  ggplot(aes(clean_n, reorder(Source, clean_n))) +
  geom_col(fill = BLUE) +
  geom_text(aes(label = scales::comma(clean_n)), hjust = -0.1, size = 3) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(x = "measurements", y = NULL) +
  theme(panel.grid.major.y = element_blank())
```

The union holds every table side by side:

```{r}
#| label: union-totals
tot |>
  transmute(Table = table, `rows at union` = clean_total) |>
  arrange(desc(`rows at union`)) |>
  kable()
```

## Harmonising the sieve cutoff

Sieved measurements are analysed on different fine fractions (finer than 63 or finer
than 20 micrometres), and one source labels the 63 micrometre cutoff as 62. The union
adds a **harmonised sieve cutoff** to every measurement: a canonical number (62 folded
to 63) and a label such as `<63um` or `<20um`, empty for bulk. This lets like-cutoff
sieved data be compared and keeps different cutoffs apart. The values themselves are
not converted between cutoffs, which is impossible without the full grain-size curve.
After harmonising, the sieved measurements are about 46,000 at `<20um` and 21,000 at
`<63um`, with only a handful at coarser cutoffs.

::: {.callout-note}
Bulk and sieved are kept as separate tracks throughout the merge: the sieve cutoff and
the bulk/sieved label are part of every later comparison, so the two are never pooled.
:::
