---
title: "Merge step 3: finalise and retention"
subtitle: "Cleaning up the keys, and how much of each source survived"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The last step turns the deduplicated union into a finished database, and lets us see
how much of each source survived.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

mdir <- "data/analysis/merge"
ret <- read_csv(file.path(mdir, "merge_retention.csv"),    show_col_types = FALSE)
tot <- read_csv(file.path(mdir, "merge_stage_totals.csv"), show_col_types = FALSE)

src_levels <- c("Mareano", "4Demon", "MUDAB", "Vannmiljø", "ICES-DOME")
BLUE <- "#4477AA"
theme_set(theme_minimal(base_size = 12))
```

## Finalising

Three things happen:

- the flagged duplicates are **removed**;
- any site, event or subsample left with no data after that removal is **dropped**
  (an ICES-DOME station whose readings were all superseded, for example);
- every temporary prefixed key is **renumbered to a clean integer**, while the source
  and the original key are kept as provenance columns for back-tracing.

The table shows the effect per table: how many rows came in from the clean databases,
how many were removed, and how many remain.

```{r}
#| label: stage-totals
tot |>
  transmute(Table = table, `from clean` = clean_total, removed,
            `in merged` = final_total, `retained %` = retained_pct) |>
  arrange(desc(`from clean`)) |>
  kable()
```

Most of the removal is measurements (the deduplication); the small drops in sites,
events and subsamples are the branches left empty once those measurements were gone.

## Retention by source

This is the key view: what fraction of each source's measurements survived into the
merged database. Four of the five sources are essentially untouched. Only ICES-DOME
falls, to about **75 percent**, because so much of it was a re-hosted copy of the
national data that the other sources already provide.

```{r}
#| label: retention
#| fig-cap: "Percentage of each source's measurements retained in the merged database."
#| fig-width: 6.5
#| fig-height: 3
ret |>
  filter(table == "measurement") |>
  mutate(Source = factor(Source, levels = src_levels)) |>
  ggplot(aes(retained_pct, Source)) +
  geom_col(fill = BLUE) +
  geom_text(aes(label = sprintf("%.1f%%", retained_pct)), hjust = -0.1, size = 3.2) +
  scale_x_continuous(limits = c(0, 108), expand = expansion(mult = c(0, 0))) +
  scale_y_discrete(limits = rev(src_levels)) +
  labs(x = "measurements retained (%)", y = NULL) +
  theme(panel.grid.major.y = element_blank())
```

The full picture across every table, per source (empty where a source has no such
table, for example 4Demon has no grain size):

```{r}
#| label: retention-table
ret |>
  mutate(Source = factor(Source, levels = src_levels),
         cell = ifelse(is.na(retained_pct), "-", sprintf("%.1f%%", retained_pct))) |>
  select(Source, table, cell) |>
  tidyr::pivot_wider(names_from = table, values_from = cell) |>
  arrange(Source) |>
  kable()
```

::: {.callout-note}
Retention here counts measurements kept versus the clean databases. A low ICES-DOME
figure is a sign of success, not data loss: those rows are duplicates of national data
that is kept from its original source. The merged database is one row per reading.
:::
