---
title: "Merge step 4: outlier flagging"
subtitle: "Marking values that sit implausibly far from their element's distribution"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Across the five sources, a small number of values are far larger or smaller than
anything else recorded for the same element. Many are registration errors: a
misplaced decimal, a unit entered as mg/kg instead of percent, a transcription slip.
The fourth step marks these so they can be reviewed, adding a soft `outlier_flag`
(`high` / `low` / NULL) to every measurement. It is a **flag, not a deletion**: a
genuinely contaminated site should survive for a person to judge, so nothing is
removed here.

This complements the physical range check carried up from the slim stage, which uses
deliberately generous fixed bounds and so misses a value that is shifted tenfold but
still under the maximum. This step instead compares each value against the **shape of
its own element's distribution**.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

mdir <- "data/analysis/merge"
smry <- read_csv(file.path(mdir, "merge_outlier_summary.csv"), show_col_types = FALSE)
hist <- read_csv(file.path(mdir, "merge_outlier_hist.csv"),    show_col_types = FALSE)
ex   <- read_csv(file.path(mdir, "merge_outlier_examples.csv"), show_col_types = FALSE)

elem_order <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN", "FE", "AL",
                "CORG", "TOC", "TOC63")
lev <- elem_order[elem_order %in% unique(smry$symbol)]
smry <- smry |> mutate(symbol = factor(symbol, levels = lev))
hist <- hist |> mutate(symbol = factor(symbol, levels = lev))

n_flag <- nrow(ex)
n_high <- sum(ex$direction == "high")
n_low  <- sum(ex$direction == "low")
frac_lab <- c(bulk = "Bulk", sieved63 = "Sieved <63 um", sieved20 = "Sieved <20 um")
```

## How a value is flagged

The check runs on the standardised values (`value_std`, mg/kg), on chemistry only
(the seven targets, the Fe/Al normalisers, and organic carbon). Grain-size
composition is bounded 0 to 100 percent and handled by its own correction, so it is
left out.

Because concentrations are strictly positive and span up to four or five orders of
magnitude, the comparison is done on **log10(value)**, and separately for each
**element and fraction**. Fraction matters: sieving concentrates metals, so a sieved
sample's normal level is higher than a bulk one, and pooling them would wrongly flag
legitimate sieved values.

Within each element-and-fraction group a value is flagged only when it is **both**:

- a statistical outlier: more than **4 robust deviations** (median absolute
  deviation) from the group median on the log scale; **and**
- at least **one order of magnitude** (tenfold) away from the group median.

The tenfold floor is what targets registration errors specifically. It keeps the
genuinely extreme values (a copper hundreds of times the typical level) while sparing
elements whose real spread is narrow, where "far in deviation units" can still be well
within a factor of ten. The effective boundary is therefore
`median +/- max(4 deviations, tenfold)`. Groups with fewer than 100 values are left
unflagged (too few to describe a distribution reliably) and rely on the physical range
check instead. Region is deliberately not used: regional differences are around
twofold, trivial next to the tenfold-plus errors this targets.

## What the distributions look like

Each panel is one element; the curves are the log10 value distribution for each
fraction, and the dashed red lines are the flagging boundary for the **bulk** track
(the sieved boundaries differ and are listed in the table below). Values outside the
dashed lines are flagged.

```{r}
#| label: distributions
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "log10(value_std) by element and fraction; dashed red = bulk flagging boundary."
bulk_thr <- smry |> filter(fraction == "bulk")

ggplot(hist, aes(bin, n, colour = fraction)) +
  geom_line(na.rm = TRUE) +
  geom_vline(data = bulk_thr, aes(xintercept = log10(thr_lo)),
             linetype = "dashed", colour = "firebrick", linewidth = 0.4) +
  geom_vline(data = bulk_thr, aes(xintercept = log10(thr_hi)),
             linetype = "dashed", colour = "firebrick", linewidth = 0.4) +
  facet_wrap(~ symbol, scales = "free") +
  scale_colour_manual(values = c(bulk = "#EE6677", sieved63 = "#4477AA",
                                 sieved20 = "#228833"),
                      labels = frac_lab, name = NULL) +
  labs(x = "log10 value (mg/kg)", y = "count") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The wide distributions (Copper, Zinc, Manganese) are flagged only far out in the
tails, where the tenfold floor is easily cleared. Narrow ones (Selenium) are governed
by the tenfold floor rather than the deviation rule, so their real tails are left
alone.

## What was flagged

In total **`r n_flag`** measurements are flagged (`r n_high` high, `r n_low` low),
about 0.3 percent of the chemistry data. The counts per element and fraction, with the
group median and the flagging boundary (all mg/kg):

```{r}
#| label: summary-table
smry |>
  arrange(symbol, fraction) |>
  transmute(Element = symbol,
            Fraction = recode(fraction, !!!frac_lab),
            N = n,
            Median = round(median_val, 1),
            `Low bound` = round(thr_lo, 1),
            `High bound` = round(thr_hi, 1),
            `Flagged high` = n_high,
            `Flagged low` = n_low) |>
  kable(format.args = list(big.mark = ","))
```

## Example flagged values

The clearest cases read as data-entry errors. Manganese at tens of thousands of mg/kg
(several percent of the sediment), organic carbon at around 100 mg/kg (a hundredth of
the typical level, a dropped decimal), copper hundreds of times its median. The
`Fold` column is the value divided by its group median.

```{r}
#| label: examples
bind_rows(
  ex |> filter(direction == "high") |> arrange(desc(fold_vs_med)) |> head(8),
  ex |> filter(direction == "low")  |> arrange(fold_vs_med)       |> head(6)
) |>
  transmute(Element = symbol,
            Fraction = recode(fraction, !!!frac_lab),
            Direction = direction,
            `Value (mg/kg)` = value_std,
            `Group median` = group_median,
            Fold = fold_vs_med,
            Source = source,
            Sea = sea_name) |>
  kable(format.args = list(big.mark = ","))
```

Some high values may be real: a heavily contaminated fjord or harbour basin can carry
metal levels far above the open-sea background. That is exactly why this stays a
review marker rather than a deletion. Downstream use can treat `outlier_flag` as a
removal candidate, weigh it together with the physical range flag and the source's own
quality flags, and decide per value.
