---
title: "Fe/Al normalisation"
subtitle: "Normalising the target metals to aluminium and iron"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Finer, more aluminosilicate-rich sediment carries more of both the lithogenic
normalisers (aluminium, iron) and the trace metals, so dividing a metal by Al (or
Fe) corrects that grain-size and mineralogical variation. A strong metal to
normaliser relationship means the metal is naturally (lithogenically) controlled,
which is the baseline the enrichment-factor / pristine determination rests on (that
cutoff is a later, merged-stage step; this page establishes the normalisation
itself).

The analysis is run **separately per source** upstream in the processing
pipeline. Each target is paired with the Al and Fe measured on
the **same subsample and same fraction** (bulk with bulk, sieved with sieved),
since normalisation is only valid within a comparable fraction. All depths are
kept, reported by the EFSA depth bands.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

ndir <- "data/analysis/normalisation"
avail <- read_csv(file.path(ndir, "normalisation_availability.csv"), show_col_types = FALSE)
corr  <- read_csv(file.path(ndir, "normalisation_correlation.csv"),  show_col_types = FALSE)
ratio <- read_csv(file.path(ndir, "normalisation_ratios.csv"),       show_col_types = FALSE)
pairs <- read_csv(file.path(ndir, "normalisation_pairs.csv"),        show_col_types = FALSE)

src_levels  <- c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon")
band_levels <- c("0-5", "5-40", ">40", "unknown")
ord_src <- function(x) factor(x, levels = src_levels)
theme_set(theme_minimal(base_size = 12))
```

## 1. Can the data be normalised?

How much of the target data carries a co-located Al or Fe value (same subsample,
same fraction), per source and depth band. Mareano measures Fe and Al with every
target; ICES-DOME, MUDAB and 4Demon cover most targets; Vannmiljø rarely does, so
it is essentially not normalisable.

```{r}
#| label: avail-table
avail |>
  mutate(Source = ord_src(Source), band = factor(band, levels = band_levels)) |>
  arrange(Source, band) |>
  select(Source, band, n_target, `% with Al` = pct_AL, `% with Fe` = pct_FE) |>
  kable(format.args = list(big.mark = ","))
```

## 2. How well does each normaliser work?

The r-squared of a log-log fit of metal against normaliser (bulk samples, at least
10 pairs). High values mean the metal is grain-size / mineralogically controlled,
so the normaliser is a sound baseline. **CO, CU and ZN** normalise well; **MN** is
weaker (it is redox-mobile); **MO and SE** are essentially uncorrelated, so
Al/Fe normalisation does not apply to them.

```{r}
#| label: strength-table
corr |>
  filter(frac_class == "bulk") |>
  select(Source, symbol, normaliser, r2) |>
  pivot_wider(names_from = normaliser, values_from = r2) |>
  transmute(Source = ord_src(Source), Element = symbol,
            `r2 (Al)` = AL, `r2 (Fe)` = FE) |>
  arrange(Source, Element) |>
  kable()
```

```{r}
#| label: strength-fig
#| fig-cap: "Log-log metal to normaliser fit strength (r-squared) in bulk samples, per source."
#| fig-width: 8
#| fig-height: 4
corr |>
  filter(frac_class == "bulk", Source %in% c("Mareano", "ICES-DOME", "MUDAB")) |>
  ggplot(aes(symbol, r2, fill = normaliser)) +
  geom_col(position = position_dodge()) +
  scale_fill_manual(values = c(AL = "#4477AA", FE = "#EE6677")) +
  facet_wrap(~ factor(Source, levels = src_levels)) +
  labs(x = NULL, y = "r-squared (log-log)", fill = "normaliser") +
  ylim(0, 1)
```

## 3. The lithogenic relationship

For the well-behaved metals, concentration scales with Al along a tight log-log
line, the natural (lithogenic) baseline. Mareano (single source, bulk, clean)
shows it most clearly; points well above the line would be candidates for
enrichment.

```{r}
#| label: scatter-fig
#| fig-cap: "Mareano bulk target concentration against aluminium (log-log, linear fit)."
#| fig-width: 8
#| fig-height: 3
pairs |>
  filter(Source == "Mareano", frac_class == "bulk",
         symbol %in% c("CO", "CU", "ZN"),
         !is.na(AL), AL > 0, value_std > 0) |>
  ggplot(aes(AL, value_std)) +
  geom_point(alpha = 0.15, size = 0.6, colour = "#4477AA") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#AA3377", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ symbol, scales = "free_y") +
  labs(x = "Al (mg/kg, log)", y = "concentration (mg/kg, log)")
```

## 4. Normalised values (metal / Al)

The median metal-to-Al ratio (bulk samples), the normalised concentration on which
a local background and enrichment factor would later be built.

```{r}
#| label: ratio-table
ratio |>
  filter(frac_class == "bulk", normaliser == "AL") |>
  transmute(Source = ord_src(Source), Element = symbol, n,
            `median Me/Al` = median, p25, p75) |>
  arrange(Source, Element) |>
  kable(format.args = list(big.mark = ","))
```

::: {.callout-note}
Normalisation applies to the grain-size-controlled metals (CO, CU, ZN, and to a
lesser extent MN). Molybdenum and selenium are not lithogenically controlled, so
their Fe/Al fits are near zero and normalisation is not appropriate for them; they
are kept in the tables for completeness. The enrichment factor and pristine cutoff
are built on these normalised values at the later merged / background stage.
:::
