---
title: "Site locations"
subtitle: "Location attributes recomputed consistently with geoenrich"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Each site's location attributes were recomputed with the
[geoenrich](https://github.com/AIQC-Hub/geoenrich) command-line tool and written
back into the clean `site` tables, so that all five sources carry **one consistent
set of location values** (derived the same way from the same reference data) rather
than each source's own. This matters for the later cross-source merge: a site off
the Norwegian coast should get the same depth, sea and distance-to-coast whether it
came from Mareano or ICES-DOME.

The step runs upstream in the processing pipeline: it exports the
site coordinates, runs geoenrich, and replaces six columns.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

sources <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon"),
  stem   = c("mareano", "vannmiljo", "ices_dome", "mudab", "4demon"))
src_levels <- sources$Source
ord_src <- function(x) factor(x, levels = src_levels)
pal <- c("Mareano" = "#4477AA", "Vannmiljø" = "#66CCEE", "ICES-DOME" = "#228833",
         "MUDAB" = "#CCBB44", "4Demon" = "#EE6677")

read_site <- function(stem) {
  con <- dbConnect(SQLite(), sprintf("data/db/%s_clean.sqlite", stem))
  on.exit(dbDisconnect(con))
  as_tibble(dbReadTable(con, "site"))
}
site <- bind_rows(lapply(seq_len(nrow(sources)), function(i)
  read_site(sources$stem[i]) |> mutate(Source = sources$Source[i])))
theme_set(theme_minimal(base_size = 12))
```

## What was recomputed

Six columns are replaced with geoenrich output; `site_id`, `latitude` and
`longitude` are untouched.

```{r}
#| label: what-table
tribble(
  ~Column,          ~Source_dataset,           ~Meaning,
  "dist_to_coast",  "GSHHG (full resolution)", "Distance to the nearest shoreline, km.",
  "depth",          "GEBCO 2024",              "Sea-floor depth, m positive-down. Points on land are set to NULL.",
  "sea_name",       "IHO Sea Areas",           "Name of the sea or ocean body.",
  "country",        "Natural Earth",           "Nearest country.",
  "country_code",   "Natural Earth",           "ISO 3166-1 alpha-3 (NOR, DEU, GBR).",
  "municipality",   "GISCO LAU (Europe)",      "Nearest municipality (Europe only)."
) |>
  rename(`Source dataset` = Source_dataset) |>
  kable()
```

## Coverage per source

```{r}
#| label: coverage-table
site |>
  group_by(Source) |>
  summarise(
    sites = n(),
    `depth known` = sum(!is.na(depth)),
    `depth NULL (land)` = sum(is.na(depth)),
    `median depth (m)` = round(median(depth, na.rm = TRUE)),
    `median coast (km)` = round(median(dist_to_coast), 1),
    countries = n_distinct(country),
    seas = n_distinct(sea_name),
    .groups = "drop") |>
  mutate(Source = ord_src(Source)) |>
  arrange(Source) |>
  kable(format.args = list(big.mark = ","))
```

Depth is NULL where a coordinate rounds onto land (fjord or estuary edges, and a
small tail of inland or mis-entered coordinates); all other depths are genuine
sea-floor values.

## Countries and seas

The dominant countries and seas per source, reflecting each source's region
(Mareano and Vannmiljø Norwegian, MUDAB German, 4Demon Belgian, ICES-DOME
international).

```{r}
#| label: country-sea-table
top_n_str <- function(df, col) {
  df |> filter(!is.na(.data[[col]])) |> count(Source, value = .data[[col]]) |>
    group_by(Source) |> slice_max(n, n = 3, with_ties = FALSE) |>
    summarise(s = paste0(value, " (", n, ")", collapse = ", "), .groups = "drop")
}
left_join(top_n_str(site, "country"), top_n_str(site, "sea_name"),
          by = "Source", suffix = c("_country", "_sea")) |>
  transmute(Source = ord_src(Source),
            `Top countries` = s_country, `Top seas` = s_sea) |>
  arrange(Source) |>
  kable()
```

## Geographic spread

```{r}
#| label: map-fig
#| fig-cap: "Site coordinates coloured by source (no basemap; longitude / latitude)."
#| fig-width: 8
#| fig-height: 5
ggplot(site, aes(longitude, latitude, colour = ord_src(Source))) +
  geom_point(alpha = 0.3, size = 0.4) +
  scale_colour_manual(values = pal) +
  coord_quickmap() +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  labs(x = "longitude", y = "latitude", colour = "Source")
```

```{r}
#| label: coast-fig
#| fig-cap: "Distance to coast by source (log scale)."
#| fig-width: 7
#| fig-height: 3
ggplot(site, aes(ord_src(Source), dist_to_coast)) +
  geom_boxplot(outlier.size = 0.3, fill = "#4477AA", alpha = 0.5) +
  scale_y_log10() +
  coord_flip() +
  labs(x = NULL, y = "distance to coast (km, log)")
```

::: {.callout-note}
Some caveats of the recomputed values. Depth is GEBCO bathymetry, so it replaces
each source's originally reported (sometimes measured) depth with a single modelled
value for comparability; land coordinates give no depth and are NULL. The
municipality is the *nearest* one from the European GISCO dataset, so an offshore
or non-European site can be assigned a distant coastal municipality. A small number
of sites sit well inland (for example an ICES coordinate on a Norwegian mountain),
which points to bad source coordinates worth flagging separately.
:::
