---
title: "Depth and distance to coast"
subtitle: "How the geographic setting relates to enrichment and the normalisers"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Do the geographic attributes **depth** and **distance to coast** affect the
normalisation? Two questions, on bulk samples, per source:

- **(B, the point)** does the **enrichment** (metal / Al) change offshore and with
  depth? A ratio that falls offshore is the coastal-pollution gradient the pristine
  background rests on: cleaner offshore sediment.
- **(A, a check)** do the **normalisers and covariates** themselves (Al, Fe,
  organic carbon, fines) vary with depth and distance to coast?

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

sdir <- "data/analysis/spatial"
enr  <- read_csv(file.path(sdir, "spatial_enrichment.csv"), show_col_types = FALSE)
nrm  <- read_csv(file.path(sdir, "spatial_normaliser.csv"), show_col_types = FALSE)
prs  <- read_csv(file.path(sdir, "spatial_pairs.csv"),      show_col_types = FALSE)

src_levels <- c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon")
ord_src <- function(x) factor(x, levels = src_levels)
theme_set(theme_minimal(base_size = 12))
```

## B. Enrichment (metal / Al) versus depth and distance to coast

Spearman correlation of the metal/Al ratio with distance to coast and with depth
(bulk samples, at least 30 per group). **Negative means the ratio falls offshore /
with depth, i.e. cleaner offshore.** Zinc shows this most strongly and
consistently, copper too; the redox-controlled Mo does not follow it.

```{r}
#| label: b-table
enr |>
  filter(normaliser == "Al") |>
  select(Source, symbol, variable, rho) |>
  pivot_wider(names_from = variable, values_from = rho) |>
  transmute(Source = ord_src(Source), Element = symbol,
            `rho vs coast` = dist_to_coast, `rho vs depth` = depth) |>
  arrange(Source, Element) |>
  kable()
```

```{r}
#| label: b-heatmap
#| fig-cap: "Metal/Al vs distance to coast (Spearman rho). Red = ratio falls offshore (coastal enrichment), blue = rises, grey = none."
#| fig-width: 6.5
#| fig-height: 3.6
enr |>
  filter(normaliser == "Al", variable == "dist_to_coast") |>
  ggplot(aes(ord_src(Source), symbol, fill = rho)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%.2f", rho), colour = abs(rho) > 0.4),
            size = 3.4, show.legend = FALSE) +
  scale_fill_gradient2(low = "#CC3311", mid = "grey92", high = "#4477AA",
                       midpoint = 0, limits = c(-0.8, 0.8), name = "rho") +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
  labs(x = NULL, y = NULL) +
  theme(panel.grid = element_blank())
```

```{r}
#| label: b-scatter
#| fig-cap: "ICES-DOME: the zinc/Al ratio against distance to coast (log y, linear fit)."
#| fig-width: 6.5
#| fig-height: 3
prs |>
  filter(Source == "ICES-DOME", symbol == "ZN",
         !is.na(ratio_AL), ratio_AL > 0, !is.na(dist_to_coast)) |>
  ggplot(aes(dist_to_coast, ratio_AL)) +
  geom_point(alpha = 0.15, size = 0.5, colour = "#4477AA") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#AA3377", se = FALSE) +
  scale_y_log10() +
  labs(x = "distance to coast (km)", y = "Zn / Al (log)")
```

The offshore decline in metal/Al (clearest for Zn and Cu) means the most enriched
sediment sits near the coast and the offshore floor is the natural background
candidate, which is what the merged-stage pristine definition will lean on.

## A. Do the normalisers themselves vary with depth and coast?

Spearman correlation of Al, Fe, organic carbon and fines with depth and distance to
coast. There is spatial structure, but its sign and strength **differ by source**
(for example Al rises offshore in MUDAB but falls in ICES-DOME), so depth and coast
are not a single, transferable effect: they are confounded with each region's grain
size, and are best handled per source.

```{r}
#| label: a-table
nrm |>
  select(Source, factor, variable, rho) |>
  pivot_wider(names_from = variable, values_from = rho) |>
  transmute(Source = ord_src(Source), Factor = factor,
            `rho vs coast` = dist_to_coast, `rho vs depth` = depth) |>
  arrange(Source, Factor) |>
  kable()
```

::: {.callout-note}
Depth and distance to coast are strongly collinear (offshore is deeper) and both
co-vary with grain size, so these are associations, not isolated effects;
separating them needs a multivariate model, a later step. Depth is modelled (GEBCO)
and absent on land-rounded sites. Vannmiljø rarely measures Al, so its metal/Al
rows rest on few samples and should be read with caution.
:::
