---
title: "Summary"
subtitle: "What the analyses found, and what to carry into the merged stage"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

This page pulls the separate analyses together in plain terms. It is meant to be
read on its own, before the detailed pages.

The trace-element data come from five sources. Before any of them can be compared,
one distinction comes first: **how the sample was measured**.

- **Bulk** means the whole sediment (or the under-2 mm part of it) was analysed.
- **Sieved** means only a fine sub-fraction was analysed, most often the material
  finer than 63 micrometres, and for the German data finer than 20 micrometres.

This matters because fine sediment concentrates metals. A sieved value is therefore
higher than a bulk value of the same mud, so the two **cannot be pooled**. We keep
them apart from the start and summarise each on its own.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

gdir <- "data/analysis/grainsize"
ndir <- "data/analysis/normalisation"
odir <- "data/analysis/organic"

frs <- read_csv(file.path(gdir, "grainsize_fraction_summary.csv"), show_col_types = FALSE)
tf  <- read_csv(file.path(gdir, "grainsize_targets_fines.csv"),    show_col_types = FALSE)
bvs <- read_csv(file.path(gdir, "grainsize_bulk_vs_sieved.csv"),   show_col_types = FALSE)
np  <- read_csv(file.path(ndir, "normalisation_pairs.csv"),        show_col_types = FALSE)
op  <- read_csv(file.path(odir, "organic_pairs.csv"),              show_col_types = FALSE)

src_levels  <- c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon")
elem_levels <- c("CO", "MN", "ZN", "CU", "MO", "SE", "I")
ord_src <- function(x) factor(x, levels = src_levels)
BULK <- "#4477AA"; SIEVED <- "#CCBB44"
theme_set(theme_minimal(base_size = 12))
```

## 1. The first split: bulk versus sieved

Two of the five sources report only bulk, one reports mostly sieved, one is almost
entirely sieved, and one carries both. The table gives the counts; the figure shows
the same shares.

```{r}
#| label: split-table
share <- frs |>
  group_by(Source, frac_class) |>
  summarise(n = sum(n), .groups = "drop") |>
  pivot_wider(names_from = frac_class, values_from = n, values_fill = 0)
if (!"sieved" %in% names(share)) share$sieved <- 0

sieve_note <- tibble(
  Source = src_levels,
  `typical sieve` = c("n/a (bulk only)", "n/a (bulk only)",
                      "<63 and <20 µm", "<20 µm", "<63 µm"))

share |>
  transmute(Source,
            `n bulk` = bulk, `n sieved` = sieved,
            `% sieved` = round(100 * sieved / (bulk + sieved))) |>
  left_join(sieve_note, by = "Source") |>
  mutate(Source = ord_src(Source)) |>
  arrange(Source) |>
  kable()
```

```{r}
#| label: split-figure
#| fig-cap: "Share of target measurements analysed on bulk (whole sample) versus sieved (fine fraction), per source."
#| fig-width: 6.5
#| fig-height: 3
frs |>
  group_by(Source, frac_class) |>
  summarise(n = sum(n), .groups = "drop") |>
  group_by(Source) |>
  mutate(share = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(share, ord_src(Source), fill = frac_class)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c(bulk = BULK, sieved = SIEVED), name = NULL) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = NULL, y = NULL) +
  theme(panel.grid.major.y = element_blank(), legend.position = "top")
```

**In short:** the Norwegian sources (Mareano, Vannmiljø) are a clean bulk world; the
international and German data (ICES-DOME, MUDAB, 4Demon) are mostly sieved, and even
the sieve size differs between them.

## 2. Bulk samples: concentration follows the sediment

In a bulk sample the number describes the whole sediment, so the concentration rises
and falls with what the sediment is made of. The analyses asked which ingredient
controls each metal. One comparable number is used throughout: the correlation of
concentration with each control, in bulk samples (positive means the metal rises
with that control).

```{r}
#| label: bulk-heatmap
#| fig-cap: "How strongly each metal tracks each control in bulk samples (median Spearman correlation across sources). Blue = rises with the control, red = falls, grey = little relation."
#| fig-width: 6.5
#| fig-height: 4
MIN_N <- 30L
control_rho <- function(df, col, label) {
  df |>
    filter(frac_class == "bulk", value_std > 0, !is.na(.data[[col]])) |>
    group_by(Source, symbol) |>
    filter(n() >= MIN_N) |>
    summarise(rho = cor(.data[[col]], value_std, method = "spearman"), .groups = "drop") |>
    group_by(symbol) |>
    summarise(control = label, rho = median(rho), .groups = "drop")
}
synth <- bind_rows(
  control_rho(tf, "fines_lt63", "Fines"),
  control_rho(np, "AL",         "Al"),
  control_rho(np, "FE",         "Fe"),
  control_rho(op, "CORG",       "Organic C")) |>
  mutate(symbol  = factor(symbol,  levels = elem_levels),
         control = factor(control, levels = c("Fines","Al","Fe","Organic C")))

ggplot(synth, aes(control, symbol, fill = rho)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%.2f", rho), colour = abs(rho) > 0.5),
            size = 3.6, show.legend = FALSE) +
  scale_fill_gradient2(low = "#CC3311", mid = "grey92", high = BULK,
                       midpoint = 0, limits = c(-0.8, 0.8), name = "correlation") +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL) +
  theme(panel.grid = element_blank())
```

Reading the rows:

- **Cobalt and manganese are lithogenic.** They rise with the fine fraction and with
  the iron / aluminium carriers. Iron or aluminium normalisation is the right
  correction for them.
- **Copper and zinc respond to everything.** They track the carriers and organic
  carbon and the fines, so either a normaliser or organic carbon can anchor a
  background.
- **Molybdenum is organic and redox driven.** It follows organic carbon but not iron
  or aluminium. Normalisation does not apply; organic carbon is its natural partner.
- **Selenium and iodine are data limited.** Selenium is weak on every control, and
  iodine has too few bulk samples to appear here at all; neither has a clear single
  control.

Two further checks on the bulk samples:

- **Location.** The metal-to-aluminium ratio falls as you move offshore and into
  deeper water, clearest for zinc and copper. The most enriched sediment sits near
  the coast, so the offshore floor is the natural-background candidate. See
  [Depth and coast](spatial.qmd).
- **Time.** There is no consistent trend across sources, and where a trend appears it
  tracks a drift in the sampling itself (later campaigns hit muddier sites), not a
  pollution history. See [Sampling year](temporal.qmd).

## 3. Sieved samples: the fine fraction is already concentrated

Sieving to the fine fraction removes the coarse, metal-poor sand, so a sieved value
is higher than the bulk value of the same sediment. The figure shows how many times
higher, per metal, for the sources that report both fractions.

```{r}
#| label: sieved-ratio
#| fig-cap: "Sieved concentration divided by bulk concentration (median), per metal and source. Above 1 (blue) means the fine fraction is enriched; below 1 (red) means it is not. Molybdenum is the exception, in line with its redox rather than grain-size control."
#| fig-width: 7
#| fig-height: 3.4
ratio <- bvs |>
  select(Source, symbol, frac_class, median) |>
  pivot_wider(names_from = frac_class, values_from = median) |>
  filter(!is.na(bulk), !is.na(sieved), bulk > 0) |>
  mutate(ratio = sieved / bulk,
         enriched = ifelse(ratio >= 1, "enriched (>=1)", "not enriched (<1)"),
         symbol = factor(symbol, levels = elem_levels))

ggplot(ratio, aes(symbol, ratio, fill = enriched)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 1, colour = "grey40", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%.1f", ratio)), vjust = -0.3, size = 3) +
  facet_wrap(~ ord_src(Source)) +
  scale_fill_manual(values = c(`enriched (>=1)` = BULK, `not enriched (<1)` = "#CC3311"),
                    name = NULL) +
  labs(x = NULL, y = "sieved / bulk") +
  theme(panel.grid.major.x = element_blank(), legend.position = "top")
```

**In short:** the grain-size-controlled metals (cobalt, copper, manganese, zinc) are
roughly two to three and a half times higher in the fine fraction, and molybdenum is
not, matching what the bulk analysis showed. Because the fine fraction is already
grain-size controlled, sieved data need far less grain-size correction. Their own
catch is that the **sieve cutoff is not the same everywhere** (63 versus 20
micrometres), so even sieved values must be brought to a common cutoff before they
are compared.

## 4. What to carry into the merged stage

The merged database will define a local natural background per element. The analyses
point to a clear recipe, and the first rule is to keep the two fractions apart.

```{r}
#| label: reco-table
tibble::tribble(
  ~Element, ~`Governing control`, ~`Bulk: how to handle`, ~`Sieved: how to handle`,
  "Co, Mn",  "Lithogenic (grain size, Fe/Al)", "Normalise by Al or Fe",          "Compare on a common sieve cutoff",
  "Cu, Zn",  "Mixed (Fe/Al and organic C)",    "Normalise by Al or Fe, or organic C", "Compare on a common sieve cutoff",
  "Mo",      "Organic / redox",                "Pair with organic carbon (not Fe/Al)", "Pair with organic carbon",
  "Se, I",   "Unclear (data limited)",         "Treat with caution, too few data", "Treat with caution, too few data"
) |>
  kable()
```

Alongside the per-element recipe:

- **Do not pool bulk and sieved.** Carry them as two tracks; a sieved value is not a
  bulk value.
- **Harmonise the sieve cutoff** among the sieved data (63 versus 20 micrometres)
  before comparing sources.
- **Use the offshore, deeper sediment as the background candidate**, per source,
  since that is where enrichment is lowest.
- **Treat time trends per source only**, and read them together with the sampling
  drift they are confounded with.

::: {.callout-note}
Every number here is a summary statistic (a Spearman correlation or a median ratio,
on bulk samples unless stated) meant for a quick read across the analyses; the fitted
models and the full breakdowns live on the individual analysis pages. Selenium and
iodine are data limited throughout.
:::
