---
title: "Synthesis"
subtitle: "Which control governs each element"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The three analyses each test one control on the target-metal concentrations: grain
size (the <63 µm fines), the Fe/Al normalisers, and organic carbon. This page ties
them together with one comparable statistic per element, to say which control
governs which metal.

For every control we take the **Spearman correlation of concentration with that
control in bulk samples**, computed per source (at least 30 bulk pairs) and then
taken as the **median across sources**, so no single source dominates. A high
positive value means the metal tracks that control.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

g <- read_csv("data/analysis/grainsize/grainsize_targets_fines.csv", show_col_types = FALSE)
n <- read_csv("data/analysis/normalisation/normalisation_pairs.csv",  show_col_types = FALSE)
o <- read_csv("data/analysis/organic/organic_pairs.csv",              show_col_types = FALSE)

MIN_N <- 30L
elem_levels    <- c("CO", "MN", "ZN", "CU", "MO", "SE", "I")
control_levels <- c("Fines", "Al", "Fe", "CORG")

# median (across sources) Spearman rho of concentration vs a control, bulk only
control_rho <- function(df, col, label) {
  df |>
    filter(frac_class == "bulk", value_std > 0, !is.na(.data[[col]])) |>
    group_by(Source, symbol) |>
    filter(n() >= MIN_N) |>
    summarise(rho = cor(.data[[col]], value_std, method = "spearman"), .groups = "drop") |>
    group_by(symbol) |>
    summarise(control = label, rho = median(rho), n_sources = n(), .groups = "drop")
}

synth <- bind_rows(
  control_rho(g, "fines_lt63", "Fines"),
  control_rho(n, "AL",         "Al"),
  control_rho(n, "FE",         "Fe"),
  control_rho(o, "CORG",       "CORG")) |>
  mutate(symbol  = factor(symbol,  levels = elem_levels),
         control = factor(control, levels = control_levels))
```

## Control strength per element

Median Spearman correlation of concentration with each control (bulk samples).
Iodine does not appear: too few bulk samples carry these controls.

```{r}
#| label: synth-table
synth |>
  select(symbol, control, rho) |>
  pivot_wider(names_from = control, values_from = rho) |>
  arrange(symbol) |>
  rename(Element = symbol) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable()
```

## Heatmap

```{r}
#| label: synth-heatmap
#| fig-cap: "Association of each metal with each control (median Spearman rho, bulk). Blue = concentration rises with the control, red = falls, grey = none."
#| fig-width: 6.5
#| fig-height: 4
ggplot(synth, aes(control, symbol, fill = rho)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%.2f", rho),
                colour = abs(rho) > 0.5), size = 3.6, show.legend = FALSE) +
  scale_fill_gradient2(low = "#CC3311", mid = "grey92", high = "#4477AA",
                       midpoint = 0, limits = c(-0.8, 0.8), name = "Spearman rho") +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "right")
```

## What governs each metal

- **Cobalt and manganese** are **lithogenic**: they track the fines and the Fe/Al
  normalisers (cobalt strongest on iron), and only weakly organic carbon. Fe/Al
  normalisation is the right correction for them.
- **Copper and zinc** respond to **everything**: strong on Fe/Al and on organic
  carbon, and positive with fines. Both a normaliser and organic carbon carry real
  signal, so either can anchor a background.
- **Molybdenum** is **organic / redox controlled**: it correlates with organic
  carbon but not with Fe or Al, and is actually negative with fines. Fe/Al
  normalisation does not apply; organic carbon is its natural covariate.
- **Selenium** is weak on every control (and sparse), and **iodine** has too few
  bulk samples to place. Neither has a clear single control here.

::: {.callout-note}
One comparable statistic (Spearman rho, bulk samples, median across sources) is used
so the three controls can be read side by side; it is an association summary, not
the fitted normalisation model on each analysis page. Selenium and iodine are
data-limited. This split (lithogenic vs organic) guides which control to use per
element when a local background is defined at the merged stage.
:::
