---
title: "Sampling year"
subtitle: "How the sampling year relates to enrichment and the normalisers"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The time analogue of the [depth and coast](spatial.qmd) page. Does the sampling
**year** affect the normalisation? Two questions, on bulk samples, per source:

- **(A, the point)** does the **enrichment** (metal / Al) trend over the years? A
  ratio that falls with year would be a real improvement signal: declining input
  relative to the lithogenic carrier.
- **(B, a check)** do the **normalisers and covariates** themselves (Al, Fe,
  organic carbon, fines) trend with year, i.e. did the sampling or method mix drift
  over time?

Unlike the coastal gradient, which pointed the same way in every source, the
temporal trend is **not consistent**: its sign differs by source, and often between
the two normalisers within a source. So year is not a transferable signal here, and
(B) shows a large part of why.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

tdir <- "data/analysis/temporal"
enr  <- read_csv(file.path(tdir, "temporal_enrichment.csv"), show_col_types = FALSE)
nrm  <- read_csv(file.path(tdir, "temporal_normaliser.csv"), show_col_types = FALSE)
prs  <- read_csv(file.path(tdir, "temporal_pairs.csv"),      show_col_types = FALSE)

src_levels <- c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon")
ord_src <- function(x) factor(x, levels = src_levels)
theme_set(theme_minimal(base_size = 12))
```

## A. Enrichment (metal / normaliser) versus sampling year

Spearman correlation of the metal/Al and metal/Fe ratios with sampling year (bulk
samples, at least 30 per group). **Negative means the ratio falls over time.** No
element points the same way across the sources, and the Al and Fe columns often
disagree within a source, so there is no shared temporal decline to read.

```{r}
#| label: a-table
enr |>
  select(Source, symbol, normaliser, rho) |>
  pivot_wider(names_from = normaliser, values_from = rho) |>
  transmute(Source = ord_src(Source), Element = symbol,
            `rho vs year (Al)` = Al, `rho vs year (Fe)` = Fe) |>
  arrange(Source, Element) |>
  kable()
```

```{r}
#| label: a-heatmap
#| fig-cap: "Metal/Al vs sampling year (Spearman rho). Blue = ratio rises over time, red = falls, grey = none. The mix of colours down each row is the point: sources disagree."
#| fig-width: 6.5
#| fig-height: 3.6
enr |>
  filter(normaliser == "Al") |>
  ggplot(aes(ord_src(Source), symbol, fill = rho)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%.2f", rho), colour = abs(rho) > 0.4),
            size = 3.4, show.legend = FALSE) +
  scale_fill_gradient2(low = "#CC3311", mid = "grey92", high = "#4477AA",
                       midpoint = 0, limits = c(-0.8, 0.8), name = "rho") +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
  labs(x = NULL, y = NULL) +
  theme(panel.grid = element_blank())
```

```{r}
#| label: a-scatter
#| fig-cap: "The zinc/Al ratio against sampling year, per source (log y, linear fit). The fitted slopes point in opposite directions, so the trend is source-specific."
#| fig-width: 7
#| fig-height: 4.5
prs |>
  filter(symbol == "ZN", !is.na(ratio_AL), ratio_AL > 0, !is.na(year)) |>
  ggplot(aes(year, ratio_AL)) +
  geom_point(alpha = 0.15, size = 0.5, colour = "#4477AA") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#AA3377", se = FALSE) +
  scale_y_log10() +
  facet_wrap(~ ord_src(Source), scales = "free_y") +
  labs(x = "sampling year", y = "Zn / Al (log)")
```

## B. Do the normalisers themselves trend with year?

Spearman correlation of Al, Fe, organic carbon and fines with sampling year. This is
where the (A) picture comes from: in several sources the **fines and organic carbon
rise with year** (ICES-DOME fines +0.47, MUDAB fines +0.37, Mareano organic carbon
+0.39), meaning later campaigns sampled muddier, more organic-rich sites. A drift in
the carrier itself moves the metal/normaliser ratio without any change in input, so
these apparent time trends are largely a change in **where and how** each source
sampled, not a pollution history.

```{r}
#| label: b-table
nrm |>
  select(Source, factor, rho) |>
  transmute(Source = ord_src(Source), Factor = factor, `rho vs year` = rho) |>
  arrange(Source, Factor) |>
  kable()
```

::: {.callout-note}
This is an unbalanced observational series, not a fixed monitoring station, so year
is confounded with the station mix, gear and lab each source used over time; these
are associations, not station trends. A real temporal signal needs station-matched
series, a later step. Mareano's year is sparse (2003 to 2021, and about a quarter of
its events carry no year), so its rows rest on fewer samples.
:::
